Is high social class always beneficial for survival? Northern Sweden 1801--2013

### 2018-03-14 marginal.Rmd

---
title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
output:
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::pdf_document2:
    citation_package: natbib
    number_sections: yes
    toc: no
    toc_depth: 2
    keep_tex: true
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
titlepage: yes
bibliography: "mybib.bib"
documentclass: article
classoption: titlepage
fontsize: 12pt
citation_package: natbib
biblio-style: "apalike"
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{graphicx}
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class. The issue is analysed according to gender and age group (adults and elderly) and in relation to the development of economic inequality. The results show that high social class is not always favourable for survival. Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern. We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology. We are adopting a marginal approach to the problem of adjust for confounders by impementing the inverse probability weighting technique."
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
source("R/oneSpell.R")
options(show.signif.stars = FALSE)
```

```{r readalldata}
load("data/skume.rda")
skume <- skume[order(skume$id, skume$enter), ]
skume$id <- paste("pop", skume$id, sep = "_")
##
skume1 <- age.window(skume, c(40, 65))  ###### OJOJOJ!!!!!!!!!!!!!!!!!!!!!!!!!!! (40, 65)
skume1 <- oneSpell(skume1)
##
skume2 <- age.window(skume, c(65, 100))
skume2 <- oneSpell(skume2)
##################################################
load("data/foli.rda")
foli$civst[is.na(foli$civst)] <- "unmarried"
foli <- foli[order(foli$id, foli$enter), ]
foli$AR <- foli$ULORSAK <- NULL
##
foli1 <- age.window(foli, c(40, 65))
foli1 <- oneSpell(foli1)
##
foli2 <- age.window(foli, c(65, 100))
foli2 <- oneSpell(foli2)
##################################################
source("CreData/R/addPer.R")
skume1 <- addPer(skume1, cuts = seq(1801, 1951, by = 25))
skume2 <- addPer(skume2, cuts = seq(1801, 1951, by = 25))
##
foli1 <- addPer(foli1, cuts = c(1976, 1986, 1996, 2005, 2014))
foli2 <- addPer(foli2, cuts = c(1976, 1986, 1996, 2005, 2014))
##################################################
vb1 <- rbind(skume1, foli1)
vb1$period <- factor(vb1$period)
##
vb2 <- rbind(skume2, foli2)
vb2$period <- factor(vb2$period)
vb <- rbind(vb1, vb2)
vb$exposure <- vb$exit - vb$enter # For statistics and figures purposes
```

@@

\begin{enumerate}
\item {\tt elite}, Hisclass 1 and 2. Higher managers and higher professionals.
\item {\tt middle}, Hisclass 3, 4, 5, and 6. Lower managers,
farm hands and lower professionals, clerical and sales personell.
\item {\tt worker}, Hisclass 7, 9, 10, 11 and 12. Workers including
farm workers.
\item {\tt farmer}, Hisclass 8. Farmers. 
\item {\tt NA}, Not Available, unknown. 
\end{enumerate}

@@

Marital status (see Figure \@ref(fig:civstskum)) is a time-dependent variable. 
The statuses used are unmarried, married and dissolved marriage
(without distinguishing between widowed and divorced). If no explicit
status is given in the sources and the partner is missing, the status has been
set to unmarried. Note that it has not been possible to identify cohabitation.

@@

axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
abline(h = 65)
box()
for (i in 2:5){
    lines(midper, resw[i, ] - refclw, col = scol[i], type = "b", pch = i, lty = i)
}
legend("bottomright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
refclm <- resm[3, ]
plot(midper, resm[1, ] - refclm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", 
     ylab = "Years", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1, xlim = c(1810, 2030))
axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
abline(h = 65)
box()
for (i in 2:5){
    lines(midper, resm[i, ] - refclm, col = scol[i], type = "b", lty = i, pch = i)
}
legend("bottomright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

Obviously, differences are getting very small as we move forward in time: Almost all forty-year-olds today survive beyond 65, regardless of social class, the exception being those who have no notation of social class.

Hovever, looking at Table \@ref(tab:cumswtab), it is obvious that the *statistical significance* is strong.

@@

```{r cumsw2, fig.cap = "Excess years of life (compared to 'worker') before 100 for a 65-year-old, by hisclass and period (experimental!).", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
##source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
mm <- resw[3, ]
plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(-4, 5), xlab = "Period", 
     ylab = "Excess years", axes = FALSE, main = "Women, 65-99")

axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
box()
for (i in 2:5){
    lines(midper, resw[i, ] - mm, col = scol[i], type = "b", lty = i, pch = i)
}

legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
mm <- resm[3, ]
plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(-4, 5), xlab = "Period", 
     ylab = "Excess years", axes = FALSE, main = "Men, 65-99")
axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
box()
for (i in 2:5){
    lines(midper, resm[i, ] - mm, col = scol[i], type = "b", lty = i, pch = i)
}
legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

>>>>

---
title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    number_sections: yes
    toc: no
    toc_depth: 2
    keep_tex: true
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
titlepage: yes
bibliography: "mybib.bib"
documentclass: article
classoption: titlepage
fontsize: 12pt
citation_package: natbib
biblio-style: "apalike"
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{graphicx}
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class. The issue is analysed according to gender and age group (adults and elderly) and in relation to the development of economic inequality. The results show that high social class is not always favourable for survival. Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern. We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology."
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warnings = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
source("R/oneSpell.R")
options(show.signif.stars = FALSE)
```

```{r readalldata}
source("R/get.file.R")
skume <- get.file("skume")
##load("data/skume.rda")
skume <- skume[order(skume$id, skume$enter), ]
skume$id <- paste("pop", skume$id, sep = "_")
##
skume1 <- age.window(skume, c(40, 65))  ###### OJOJOJ!!!!!!!!!!!!!!!!!!!!!!!!!!! (40, 65)
skume1 <- oneSpell(skume1)
##
skume2 <- age.window(skume, c(65, 100))
skume2 <- oneSpell(skume2)
##################################################
foli <- get.file("foli")
##load("data/foli.rda")
foli$civst[is.na(foli$civst)] <- "unmarried"
foli <- foli[order(foli$id, foli$enter), ]
foli$AR <- foli$ULORSAK <- NULL
##
foli1 <- age.window(foli, c(40, 65))
foli1 <- oneSpell(foli1)
##
foli2 <- age.window(foli, c(65, 100))
foli2 <- oneSpell(foli2)
##################################################
source("CreData/R/addPer.R")
skume1 <- addPer(skume1, cuts = seq(1801, 1951, by = 25))
skume2 <- addPer(skume2, cuts = seq(1801, 1951, by = 25))
##
foli1 <- addPer(foli1, cuts = c(1976, 1986, 1996, 2005, 2014))
foli2 <- addPer(foli2, cuts = c(1976, 1986, 1996, 2005, 2014))
##################################################
vb1 <- rbind(skume1, foli1)
vb1$period <- factor(vb1$period)
##
vb2 <- rbind(skume2, foli2)
vb2$period <- factor(vb2$period)
vb <- rbind(vb1, vb2)
vb$exposure <- vb$exit - vb$enter # For statistics and figures purposes
```

@@

\begin{enumerate}
\item {\tt elite}, Hisclass 1 and 2. Higher managers and higher professionals.
\item {\tt middle}, Hisclass 3, 4, 5, and 6. Lower managers,
farm hands and lower professionals, clerical and sales personell.
\item {\tt worker}, Hisclass 7, 9, 10, 11 and 12. Workers including
farm workers.
\item {\tt farmer}, Hisclass 8. Farmers. 
\item {\tt none}, not given, unknown. 
\end{enumerate}

@@

Marital status (see Figure \@ref(fig:civstskum)) is a time-fixed covariate.
The statuses used are unmarried, married and dissolved marriage
(without distinguishing between widowed and divorced). If no explicit
status is given in the sources and the partner is missing, the status has been
set to unmarried. Note that it has not been possible to identify cohabitation.

@@

axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 65)
box()
for (i in 2:5){
    lines(midper, resw[i, ] - refclw, col = scol[i], type = "b", pch = i, lty = i)
}
legend("bottomright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
refclm <- resm[3, ]
plot(midper, resm[1, ] - refclm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", 
     ylab = "Years", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1, xlim = c(1810, 2030))
axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1,)
abline(h = 65)
box()
for (i in 2:5){
    lines(midper, resm[i, ] - refclm, col = scol[i], type = "b", lty = i, pch = i)
}
legend("bottomright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

Obviously, differences are getting very small as we move forward in time: Almost all forty-year-olds today survive beyond 65, regardless of social class, the exception being those who have no notation of social class. In Sweden of the year 2014, the probability of surviving age 65 for a forty-year-old man is 91 per cent  (a period measure). The corresponding figure for a forty-year-old woman is 94 per cent. More interesting is to look at the late differences in the age group 65--100.

Hovever, from Table \@ref(tab:cumswtab) it is obvious that the *statistical significance* is strong.

@@

```{r cumsw2, fig.cap = "Excess years of life (compared to 'worker') before 100 for a 65-year-old, by hisclass and period (experimental!).", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
##source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
mm <- resw[3, ]
plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 5), xlab = "Period", 
     ylab = "Excess years", axes = FALSE, main = "Women, 65-99")

axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1)
box()
for (i in 2:5){
    lines(midper, resw[i, ] - mm, col = scol[i], type = "b", lty = i, pch = i)
}

legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
mm <- resm[3, ]
plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 5), xlab = "Period", 
     ylab = "Excess years", axes = FALSE, main = "Men, 65-99")
axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1)
box()
for (i in 2:5){
    lines(midper, resm[i, ] - mm, col = scol[i], type = "b", lty = i, pch = i)
}
legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```
